### 2021-08-30 submission/response_to_reviewers.md

---
output:
  pdf_document: default
  html_document: default
header-includes:
    - \usepackage{setspace}\doublespacing  
---
### Reviewer #1 (Comments for the Author):

**In this manuscript, Tomkovich et al. examined experiments to show how PEG impacts mice susceptibility to C. difficile. They found PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remained colonized for up to 30 days post-challenge. In contrast, 1-day PEG treated mice were transiently colonized, clearing C. difficile within 7 days post-challenge. They concluded that osmotic laxatives disrupt colonization resistance to C. difficile, and prevent clearance among mice already colonized with C. difficile.**

**The study is very interesting and the data is strong to support their findings. I only have several minor comments following:**

**Line 47-49 Please clarify and make it more clear to led to your hypothesis.**

> We have revised to:
"These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea. For example, laxatives may disrupt colonization resistance to *C. difficile*."

**Line 720 Please clear Asterisks mean. Authors noticed it as the difference between groups, it's not clear which two groups (inter-group at the same time point or same group compared to day 0) were compared and showed a significant difference, please clear it for Fig. 1B and 1C.**

> We have rephrased to the following:
"Asterisks indicate the specific time points on the x-axis where the weight change or CFU/g was significantly different (*P* < 0.05) between the 3-4 groups indicated on the plot by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points."

**Did the author detect PEG concentration from stools along with experiments? Or any references can be referred to PEG metabolization and residence time in the gut?**

> We did not measure PEG concentrations as previous studies report PEG is nonabsorbable and not metabolized by bacteria (Hammer et al. JCI 1989; Di Palma et al. The American Journal of Gastroenterology 2002).

**Line 148 Did author detect the toxin concentration from fecal samples? It's interesting that more C. difficile CFU was detected and high weight loss in PEG treated group, while no significant difference in the gut.**

> We did not measure toxin concentration because we observed no histological differences between PEG and clindamycin treated mice. We agree it is an interesting observation, but suspect this may be partially due to the strain we used (*C. difficile* 630), and bring up this possibility in the discussion (lines 315-320).

**Line 176: Is it possible to include the histopathology figures in Fig. 3D?**

> *C. difficile* 630 typically causes mild histopathology and pictures are included in Theriot et al. Gut Microbes 2011 which is referred to in the methods section (lines 441-442). Because of figure size constraints and since histopathology was mild for all groups, we decided not to include histopatholgy pictures.

**Fig. 3D Label symbols of Clind. Without infection and Clind, After 5-day, 5-day PEG without infection and 5-day PEG are the same. Please correct.**

> The open and closed circles differentiate mice that were infected with *C. difficile* versus mice that were not. However, there were 2 sets of open circles that were the same color for the After 5-day PEG (Day 0) and 5-day PEG without infection groups, so we have added an additional shape to differentiate the Day 0 group in the Fig. 3D legend. Additionally, we have combined the scales to make the differences between groups more clear (Fig. 3D-E).

**Line 276-277 Could you notice which result Figure (Maybe Fig. 3C) you are discussed to help the reader to follow?**

> We have added a reference to Fig. 7B in the sentence.

**Line 284-285 Spores persistence in cell should be also noticed in this part as it can also contribute to C. difficile persistence. Ref: Castro-Córdova P, Mora-Uribe P, Reyes-Ramírez R, et al. Nature communications, 2021.**

> We have modified the sentence to the following:
"One possible explanation for the prolonged *C. difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment either within host cells (Castro-Córdova et al. Nature Communications 2021) or together with other bacteria."

**Line 256-358 So many parts for discussion, maybe can concentrate and minimize paragraphs.**

> The manuscript is within mSphere's 5000 word limit (excluding the Materials and Methods, References, and figure legends) for research articles. Given all the available literature on the microbiota, laxatives, and *C. difficile*, we wanted place our findings in context, which resulted in a longer discussion.

**Line 390 Stool was collected as fresh one or obtained from cages? How did you process your stool samples? Plating every day for CFU check or stock -80 {degree sign}C and did CFU check at the end of the experiment?**

> We have added the following sentences to clarify our stool processing methods:
"Fresh stool samples were collected from each mouse and split into two separate tubes. One tube was transferred to an anaerobic chamber the same day the sample was collected to quantify *C. difficile*, while the other tube was snap frozen in liquid nitrogen and stored in the -80&deg;C for 16S rRNA sequencing."

**Specific comments:**
**Over-referred in this manuscript (80 references were included).**

> There is no limit on the number of references for mSphere research articles.

**Typos**
**Line 4 Alter-alters**
**Line 59 Challenge-the challenge**
**Line 181 Signifant---Significant**
**Line 267 received**
**Line 312 Colonization**

> We thank the reviewer for identifying these typos and have corrected all except for line 4. Alter was used because we are stating multiple medications alter the microbiota and just giving PEG as a specific example.

### Reviewer #2 (Comments for the Author):

**As expected from this group, this is a well conducted study that is timely relevant for CDI. In this work authors address an intriguing question related with how laxative treatment affects the susceptibility of the host to C. difficile infections. Authors provide a comprehensive and very detailed overview of the impact of the osmotic laxative, PEG, to shifts in fecal and mucosal microbiota, and how PEG impacted the susceptibility of mice to infection and subsequent clearance. This work is of great public relevance and complements several other studies published on this topic.**

**I only have minor comments for the authors.**

**Specific comments:**
**Line 324: not clear please revise**

> We have revised the sentence to the following:
"The extent to which laxatives disrupt *C. difficile* colonization resistance in human patients is unclear based on the current literature."

**Figure 5: is there any reason why FMT was not used alone as a control in this experiment? There is data on how FMT impacts the microbiota composition in Figure 6, but not in the C. difficile infection experiment. Could the authors clarify this?**

> We did not use FMT alone in this experiment because clindamycin treated mice typically clear *C. difficile* within 10 days without any additional treatments such as an FMT.

**Figure 8: Please revise the narrative of the figure legend for consistencies in language.**

> We have modified the legend to:
"The gut microbiota of our lab's C57Bl/6 mice is resistant to *C. difficle* but treatment with either clindamycin or the osmotic laxative, PEG 3350, renders the mice susceptible to *C. difficile* colonization. Recovery of colonization resistance in clindamycin-treated mice is relatively straightforward and the mice clear *C.difficile* within 10 days post-challenge. However, recovery of colonization resistance was delayed for mice that received either a 5-day PEG pre-treatment or a 1-day PEG post-*C. difficile* challenge treatment. We found increased relative abundances of *Porphyromonadaceae* and *Lachnospiraceae* were associated with *C. difficle* clearance, while increased relative abundances of *Enterobacteriaceae* and *Bacteroides* were associated with prolonged *C. difficle* colonization."

### 2021-08-30 submission/manuscript.Rmd
Add edits and response to reviewers for revision

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative samples from patients were separated into two groups based on stool consistency, there were similar microbiota features between samples from CDI patients and *C. difficile* negative patients with diarrhea compared to non-diarrheal samples that were *C. difficile* negative [@Schubert2014]. The similar community features between CDI patients and patients with diarrhea included low alpha diversity and only 6 bacterial taxa with higher relative abundances in communities from CDI patients. These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.

Depending on the dose administered, osmotic laxatives can lead to diarrhea and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

Because there were differences in the mucosal microbiota, including detectable *C. difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C. difficile* challenge by evaluating cecum and colon histopathology [@Reeves2011]. However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although there was a slight difference in the histopathology score of the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5 to 2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

## Discussion
While the disruptive effect of antibiotics on *C. difficile* colonization resistance is well established, the extent to which other drugs such as laxatives disrupt colonization resistance was unclear. By following mice treated with an osmotic laxative over time, we found that a 5-day PEG treatment before challenge resulted in prolonged *C. difficile* colonization, while a 1-day PEG treatment resulted in transient colonization without the use of antibiotics. The differences in *C. difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in the degree to which treatments disrupted the microbiota. Additionally, the intestinal communities of 5-day PEG treated mice did not regain colonization resistance after a 10-day recovery period. In contrast to the other 5-day PEG treatment groups, *C. difficile* was not immediately detectable in the stools of most of the mice in the 10-day recovery group. We also examined the impact of PEG treatment after *C. difficile* challenge. In opposition to the hypothesis suggested by the literature, we found that PEG treatment prolonged colonization relative to mice that only recieved clindamycin treatment. We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C. difficile* colonization (Fig. 8). Overall, our results demonstrated that osmotic laxative treatment alone rendered mice susceptible to *C. difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after challenge.

In addition to altering composition, laxative treatment may alter microbiota-produced metabolites. A previous study demonstrated that a 5-day treatment of 10% PEG depleted acetate and butyrate and increased succinate compared to untreated mice [@Ferreyra2014]. While we did not perform metabolomic analysis, we did see bacteria known to produce beneficial metabolites were depleted in mice that cleared *C. difficile* compared to mice with prolonged colonization. For example, *Oscillibacter valericigenes* can produce the SCFA valerate [@Iino2007], and separate studies demonstrated valerate is depleted after clindamycin treatment and inhibited *C. difficile* growth *in vitro* and in C57BL/6 mice [@Jump2014; @McDonald2018]. Similarly, *Acetatifactor* can produce acetate and butyrate [@Pfeiffer2012], SCFAs that are decreased in mice with prolonged *C. difficile* infection after antibiotic treatment [@Lawley2012]. Thus protective bacteria and their metabolites could be depleted by osmotive laxative treatment depending on the timing and duration of treatment.

One possible explanation for the prolonged *C. difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment. In fact, it has been hypothesized that *C. difficile* biofilms may serve as reservoirs for recurrent infections [@Frost2021] and *C. difficile* biofilms in the mucus layer were recently identified in patients as aggregates with *Fusobacterium nucleatum* [@Engevik2021]. There was an interesting pattern of increased *Enterobacteriaceae*, *Bacteroides*, and *C. difficile* in both the stool and mucosal communities of PEG-treated mice suggesting a potential synergy. *Bacteroides* has the potential to degrade mucus and the osmotic laxative may have allowed *Bacteroides* to colonize the mucosal niche by degrading mucin glycans with glycosyl hydrolases that are absent in *C. difficile* [@Engevik2020]. *Bacteroides* persistence in the mucosal tissue might also have helped *Enterobacteriaceae* to colonize the region, as a synergy between mucus-degrading *B. fragilis* and *E. coli* has previously been described [@Dejea2018]. A separate study demonstrated *C. difficile* was present in the outer mucus layer and associated with *Enterobacteriaceae* and *Bacteroidaceae* using fluorescent in situ hybridization (FISH) staining [@Semenyuk2015]. However, protective roles for *Bacteroides* have also been demonstrated. For example, *B. fragilis* prevented CDI morbidity in a mouse model and inhibited *C. difficile* adherence *in vitro* [@Deng2018]. In coculture experiments, *B. longum* decreased *C. difficile* biofilm formation while *B. thetaiotamicron* enhanced biofilm formation [@Normington2021] and *B. dorei* reduced *C. difficile* growth in a 9-species community *in vitro* [@Hassall2021]. Therefore, whether *Bacteroides* is detrimental or beneficial in the context of *C. difficile* infection or colonization is still unclear, but the niche and interactions with other bacteria may contribute.

*Akkermansia* is also a mucin degrader with potentially beneficial or detrimental roles depending on context in other diseases [@Cirstea2018; @BorgesCanha2015]. In our study, the relative abundance of *Akkermansia* shifted over time between groups of mice that either cleared *C. difficile* or had prolonged colonization. In the stool it was initially increased in mice that cleared *C. difficile*, but shifted after 5-days post-challenge so that it was increased in mice that had prolonged colonization. In the context of CDIs, some studies suggest a protective role [@Pereira2020; @Rodriguez2016], while others suggest a detrimental role because *Akkermansia* was positively correlated with *C. difficile* [@Sangster2016; @Vakili2020a; @Vakili2020b; @Li2019]. Because the relative abundance of *Akkermansia* was dynamic in our study, it is unclear whether *Akkermansia* helps with clearance of *C. difficile* or allows it to persist. A better understanding how *C. difficile* interacts with the mucosal microbiota may lead to insights into CDIs, asymptomatic *C. difficile* carriage, and colonizatiion resistance.

@@

It is more difficult to interpret what are findings mean in the context of *C. difficile* colonization resistance in human patients. The main difficulty being that most hospitals recommend not performing *C. difficile* testing if the patient is currently taking a laxative. This recommendation is in accordance with the Infectious Diseases Society of America and Society for Healthcare Epidemiology of America guidelines [@McDonald2018b]. The rationale behind the recommendation is that patients taking laxatives may be asymptomatically colonized with *C. difficile*, resulting in unnecessary antibiotic treatment [@Ahmad2017; @Bilinskaya2018; @Cook2020]. Furthermore, some studies identified laxatives as a risk factor for developing CDIs or recurrent CDIs [@Appaneal2019; @Dubberke2011; @Obritsch2010] and a recent study found the proportion of severe CDIs was similar between patients taking and not taking laxatives [@White2019]. However, there have also been some studies that suggest laxatives are not a risk factor for developing CDIs [@Poirier2019; @Huang2014]. Although, it is unclear whether laxatives impact CDI susceptibility in human paitents, it is clear that laxatives also have a transient impact on the human microbiota [@Nagata2019; @Gorkiewicz2013; @Shobar2016; @Drago2016; @Jalanka2014]. Additional studies to examine the relationship between laxatives, *C. difficile* colonization, and CDIs are warranted.

@@

***C. difficile* challenge model.** Mice were challenged with 25 microliters containing 10^5^ *C. difficile* 630 spores, except for 1 experiment where the concentration was 10^3^ (Fig. 5A). All mock challenged mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper pipette was used to administer the same challenge dose to mice via oral gavage. Mice were weighed daily throughout the experiment and stool was collected for quantifying *C. difficile* CFU and 16S rRNA gene sequencing. There were two groups of mice that received either a PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal microbiota transplant was prepared with stool samples collected from the mice in the experiment prior to the start of any treatments. The stool samples were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS and glycerol was added to make a 15% glycerol solution. The solution was then aliquoted into tubes and stored at -80&deg;C until the day of the gavage. An aliquot of both the FMT and PBS solutions were also set aside in the -80&deg;C for 16S rRNA gene sequencing. The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The supernatant was then transferred to a separate tube to prevent the gavage needle from clogging with debris during gavage. The PBS solution that was administered to the other group was also 15% glycerol. Each mouse was administered 100 microliters of either the FMT or PBS solution via gavage. When we refer to mice that cleared *C. difficile*, we mean that no *C. difficile* was detected in the first serial dilution (limit of detection: 100 CFU). In some experiments, we collected tissues for 16SrRNA gene sequencing, histopathology, or both. For 16S rRNA gene sequencing, we collected small snips of cecum, proximal colon, and distal colon tissues in microcentrifuge tubes, snap froze in liquid nitrogen, and stored at -80&deg;C. For histopathology, cecum and colon tissues were placed into separate cassettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.

@@

\includegraphics{figure_1.pdf}

**Figure 1. 5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.** A. Setup of the experimental time line for experiments with 5-day PEG treated mice consisting of 4 treatment groups. 1. Clindamycin was administered at 10 mg/kg by intraperitoneal injection. 2. 15% PEG 3350 was administered in the drinking water for five days. 3. 5-day PEG plus clindamycin treatment. 4. 5-day PEG plus 10-day recovery treatment. All treatment groups were then challenged with 10^5^ *C. difficile* 630 spores. A subset of mice were euthanized on either 4 or 6 days post-challenge and tissues were collected for histopathology analysis, the remaining mice were followed through 20 or 30 days post-challenge. B. Weight change from baseline weight in groups after treatment with PEG and/or clindamycin, followed by *C. difficile* challenge. C. *C. difficile* CFU/gram stool measured over time via serial dilutions(N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per time point). The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths. For B-C, lines represent the median for each treatment group and circles represent samples from individual mice. Asterisks indicate time points where the weight change or CFU/g was significantly different (*P* < 0.05) between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points. The data presented are from a total of `r five_day_total_exp` separate experiments.
\newpage

\includegraphics{figure_2.pdf}
**Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. Principal Coordinate analysis (PCoA) of Bray-Curtis distances from stool samples collected throughout the experiment. Each circle represents a sample from an individual mouse and the transparency of the symbol corresponds to the day post-challenge. See Data Set S1, sheet 1 for PERMANOVA results. B. Bray-Curtis distances of stool samples collected on either day 10 or 30 post-challenge relative to the baseline sample collected for each mouse (before any drug treatments were administered). The symbols represent samples from individual mice and the line indicates the median value for each treatment group. C. Shannon diversity in stool communities over time. The line indicates the median value for each treatment group (Data Set S1, sheet 2). D. 14 of the `r f_d_group_total_genera` genera affected by PEG treatment  (Data Set S1, sheet 3). The symbols represent the median relative abundance for a treatment group at either baseline (open circle) or 1-day post treatment (closed circle). Relative abundance data from paired baseline and 1-day post treatment stool sampes from the 5-day PEG and 5-day PEG plus 10-day recovery groups were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera. The clindamycin and 5-day PEG plus clindamycin treatment groups are shown on the plot for comparison. E. 6 of the `r f_d_group_multi_day` genera that were significantly different between the treatment groups over multiple time points (see Data Set S1, sheet 4 for complete list). The 5-day PEG plus clindamycin treatment group was only followed through 6-days post-challenge. Differences between treatment groups were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified genera (\*\, *P* < 0.05). The gray vertical line (D) and horizontal vertical lines (E) indicate the limit of detection.
\newpage

>>>>

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative samples from patients were separated into two groups based on stool consistency, there were similar microbiota features between samples from CDI patients and *C. difficile* negative patients with diarrhea compared to non-diarrheal samples that were *C. difficile* negative [@Schubert2014]. The similar community features between CDI patients and patients with diarrhea included low alpha diversity and only 6 bacterial taxa with higher relative abundances in communities from CDI patients. These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea. For example, laxatives may disrupt colonization resistance to *C. difficile*.

Depending on the dose administered, osmotic laxatives can lead to diarrhea and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after the challenge is unclear.

@@

Because there were differences in the mucosal microbiota, including detectable *C. difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C. difficile* challenge by evaluating cecum and colon histopathology [@Reeves2011]. However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although there was a slight difference in the histopathology score of the colon between PEG and clindamycin-treated mice, there was not a significant difference in the cecum and the overall score was relatively low (1.5 to 2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

## Discussion
While the disruptive effect of antibiotics on *C. difficile* colonization resistance is well established, the extent to which other drugs such as laxatives disrupt colonization resistance was unclear. By following mice treated with an osmotic laxative over time, we found that a 5-day PEG treatment before challenge resulted in prolonged *C. difficile* colonization, while a 1-day PEG treatment resulted in transient colonization without the use of antibiotics. The differences in *C. difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in the degree to which treatments disrupted the microbiota. Additionally, the intestinal communities of 5-day PEG treated mice did not regain colonization resistance after a 10-day recovery period. In contrast to the other 5-day PEG treatment groups, *C. difficile* was not immediately detectable in the stools of most of the mice in the 10-day recovery group. We also examined the impact of PEG treatment after *C. difficile* challenge. In opposition to the hypothesis suggested by the literature, we found that PEG treatment prolonged colonization relative to mice that only received clindamycin treatment. We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C. difficile* colonization (Fig. 8). Overall, our results demonstrated that osmotic laxative treatment alone rendered mice susceptible to *C. difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after challenge.

In addition to altering composition, laxative treatment may alter microbiota-produced metabolites. A previous study demonstrated that a 5-day treatment of 10% PEG depleted acetate and butyrate and increased succinate compared to untreated mice [@Ferreyra2014]. While we did not perform metabolomic analysis, we did see bacteria known to produce beneficial metabolites were depleted in mice that cleared *C. difficile* compared to mice with prolonged colonization (Fig. 7B). For example, *Oscillibacter valericigenes* can produce the SCFA valerate [@Iino2007], and separate studies demonstrated valerate is depleted after clindamycin treatment and inhibited *C. difficile* growth *in vitro* and in C57BL/6 mice [@Jump2014; @McDonald2018]. Similarly, *Acetatifactor* can produce acetate and butyrate [@Pfeiffer2012], SCFAs that are decreased in mice with prolonged *C. difficile* infection after antibiotic treatment [@Lawley2012]. Thus protective bacteria and their metabolites could be depleted by osmotive laxative treatment depending on the timing and duration of treatment.

One possible explanation for the prolonged *C. difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment either within host cells [@CastroCrdova2021] or together with other bacteria. In fact, it has been hypothesized that *C. difficile* biofilms may serve as reservoirs for recurrent infections [@Frost2021] and *C. difficile* biofilms in the mucus layer were recently identified in patients as aggregates with *Fusobacterium nucleatum* [@Engevik2021]. There was an interesting pattern of increased *Enterobacteriaceae*, *Bacteroides*, and *C. difficile* in both the stool and mucosal communities of PEG-treated mice suggesting a potential synergy. *Bacteroides* has the potential to degrade mucus and the osmotic laxative may have allowed *Bacteroides* to colonize the mucosal niche by degrading mucin glycans with glycosyl hydrolases that are absent in *C. difficile* [@Engevik2020]. *Bacteroides* persistence in the mucosal tissue might also have helped *Enterobacteriaceae* to colonize the region, as a synergy between mucus-degrading *B. fragilis* and *E. coli* has previously been described [@Dejea2018]. A separate study demonstrated *C. difficile* was present in the outer mucus layer and associated with *Enterobacteriaceae* and *Bacteroidaceae* using fluorescent in situ hybridization (FISH) staining [@Semenyuk2015]. However, protective roles for *Bacteroides* have also been demonstrated. For example, *B. fragilis* prevented CDI morbidity in a mouse model and inhibited *C. difficile* adherence *in vitro* [@Deng2018]. In coculture experiments, *B. longum* decreased *C. difficile* biofilm formation while *B. thetaiotamicron* enhanced biofilm formation [@Normington2021] and *B. dorei* reduced *C. difficile* growth in a 9-species community *in vitro* [@Hassall2021]. Therefore, whether *Bacteroides* is detrimental or beneficial in the context of *C. difficile* infection or colonization is still unclear, but the niche and interactions with other bacteria may contribute.

*Akkermansia* is also a mucin degrader with potentially beneficial or detrimental roles depending on context in other diseases [@Cirstea2018; @BorgesCanha2015]. In our study, the relative abundance of *Akkermansia* shifted over time between groups of mice that either cleared *C. difficile* or had prolonged colonization. In the stool it was initially increased in mice that cleared *C. difficile*, but shifted after 5-days post-challenge so that it was increased in mice that had prolonged colonization. In the context of CDIs, some studies suggest a protective role [@Pereira2020; @Rodriguez2016], while others suggest a detrimental role because *Akkermansia* was positively correlated with *C. difficile* [@Sangster2016; @Vakili2020a; @Vakili2020b; @Li2019]. Because the relative abundance of *Akkermansia* was dynamic in our study, it is unclear whether *Akkermansia* helps with clearance of *C. difficile* or allows it to persist. A better understanding how *C. difficile* interacts with the mucosal microbiota may lead to insights into CDIs, asymptomatic *C. difficile* carriage, and colonization resistance.

@@

The extent to which laxatives disrupt *C. difficile* colonization resistance in human patients is unclear based on the current literature. The main difficulty being that most hospitals recommend not performing *C. difficile* testing if the patient is currently taking a laxative. This recommendation is in accordance with the Infectious Diseases Society of America and Society for Healthcare Epidemiology of America guidelines [@McDonald2018b]. The rationale behind the recommendation is that patients taking laxatives may be asymptomatically colonized with *C. difficile*, resulting in unnecessary antibiotic treatment [@Ahmad2017; @Bilinskaya2018; @Cook2020]. Furthermore, some studies identified laxatives as a risk factor for developing CDIs or recurrent CDIs [@Appaneal2019; @Dubberke2011; @Obritsch2010] and a recent study found the proportion of severe CDIs was similar between patients taking and not taking laxatives [@White2019]. However, there have also been some studies that suggest laxatives are not a risk factor for developing CDIs [@Poirier2019; @Huang2014]. Although, it is unclear whether laxatives impact CDI susceptibility in human paitents, it is clear that laxatives also have a transient impact on the human microbiota [@Nagata2019; @Gorkiewicz2013; @Shobar2016; @Drago2016; @Jalanka2014]. Additional studies to examine the relationship between laxatives, *C. difficile* colonization, and CDIs are warranted.

@@

***C. difficile* challenge model.** Mice were challenged with 25 microliters containing 10^5^ *C. difficile* 630 spores, except for 1 experiment where the concentration was 10^3^ (Fig. 5A). All mock challenged mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper pipette was used to administer the same challenge dose to mice via oral gavage. Mice were weighed daily throughout the experiment and stool was collected for quantifying *C. difficile* CFU and 16S rRNA gene sequencing. Fresh stool samples were collected from each mouse and split into two separate tubes. One tube was transferred to an anaerobic chamber the same day the sample was collected to quantify *C. difficile*, while the other tube was snap frozen in liquid nitrogen and stored in the -80&deg;C for 16S rRNA sequencing. There were two groups of mice that received either a PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal microbiota transplant was prepared with stool samples collected from the mice in the experiment prior to the start of any treatments. The stool samples were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS and glycerol was added to make a 15% glycerol solution. The solution was then aliquoted into tubes and stored at -80&deg;C until the day of the gavage. An aliquot of both the FMT and PBS solutions were also set aside in the -80&deg;C for 16S rRNA gene sequencing. The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The supernatant was then transferred to a separate tube to prevent the gavage needle from clogging with debris during gavage. The PBS solution that was administered to the other group was also 15% glycerol. Each mouse was administered 100 microliters of either the FMT or PBS solution via gavage. When we refer to mice that cleared *C. difficile*, we mean that no *C. difficile* was detected in the first serial dilution (limit of detection: 100 CFU). In some experiments, we collected tissues for 16SrRNA gene sequencing, histopathology, or both. For 16S rRNA gene sequencing, we collected small snips of cecum, proximal colon, and distal colon tissues in microcentrifuge tubes, snap froze in liquid nitrogen, and stored at -80&deg;C. For histopathology, cecum and colon tissues were placed into separate cassettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.

@@

**Figure 1. 5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.** A. Setup of the experimental time line for experiments with 5-day PEG treated mice consisting of 4 treatment groups. 1. Clindamycin was administered at 10 mg/kg by intraperitoneal injection. 2. 15% PEG 3350 was administered in the drinking water for five days. 3. 5-day PEG plus clindamycin treatment. 4. 5-day PEG plus 10-day recovery treatment. All treatment groups were then challenged with 10^5^ *C. difficile* 630 spores. A subset of mice were euthanized on either 4 or 6 days post-challenge and tissues were collected for histopathology analysis, the remaining mice were followed through 20 or 30 days post-challenge. B. Weight change from baseline weight in groups after treatment with PEG and/or clindamycin, followed by *C. difficile* challenge. C. *C. difficile* CFU/gram stool measured over time via serial dilutions(N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per time point). The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths. For B-C, lines represent the median for each treatment group and circles represent samples from individual mice. Asterisks indicate the specific time points on the x-axis where the weight change or CFU/g was significantly different (*P* < 0.05) between the 3-4 groups indicated on the plot by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points. The data presented are from a total of `r five_day_total_exp` separate experiments.
\newpage

**Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. Principal Coordinate analysis (PCoA) of Bray-Curtis distances from stool samples collected throughout the experiment. Each circle represents a sample from an individual mouse and the transparency of the symbol corresponds to the day post-challenge. See Data Set S1, sheet 1 for PERMANOVA results. B. Bray-Curtis distances of stool samples collected on either day 10 or 30 post-challenge relative to the baseline sample collected for each mouse (before any drug treatments were administered). The symbols represent samples from individual mice and the line indicates the median value for each treatment group. C. Shannon diversity in stool communities over time. The line indicates the median value for each treatment group (Data Set S1, sheet 2). D. 14 of the `r f_d_group_total_genera` genera affected by PEG treatment  (Data Set S1, sheet 3). The symbols represent the median relative abundance for a treatment group at either baseline (open circle) or 1-day post treatment (closed circle). Relative abundance data from paired baseline and 1-day post treatment stool sampes from the 5-day PEG and 5-day PEG plus 10-day recovery groups were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera. The clindamycin and 5-day PEG plus clindamycin treatment groups are shown on the plot for comparison. E. 6 of the `r f_d_group_multi_day` genera that were significantly different between the treatment groups over multiple time points (see Data Set S1, sheet 4 for complete list). The 5-day PEG plus clindamycin treatment group was only followed through 6-days post-challenge. Differences between treatment groups were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified genera (\*\, *P* < 0.05). The gray vertical line (D) and horizontal vertical lines (E) indicate the limit of detection.
\newpage
